---
title: "Seminar 1 notes"
date: 10.10.23
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

### **Seminar 1 10.10.23 notes**

#### Tips til Repeat

Lage en "package" med datasettene våre som alle kan laste ned

#### ggplot

Data --\> asthetics --\> geometric objects --\> facets --\> statistical transformations --\> coordinates --\> theme and annotations

#### How to make "senket skrift"

VO~2max~

#### Fotnote

This is a \[\^footnote\]

\[\^footnote\]

### Other tips

In quarto you can make html, pdf and docx from the same quatro-file. Make html-files first, then docx or pdf

#### Code chuncks

ctrl+alt+i = new code chunck

#### Pipes

\|\> or %\>% ?

Double :: means search for the dataset inside the package

```{r}

library(tidyverse)
library(exscidata)

dplyr::starwars

# Base R forward pipe operator
exscidata::dxadata |> # Smart to define that I am searching for this dataset in this package
  select(participant)
  
# Magritter forward pipe operator
exscidata::dxadata %>% 
  select(participant)
  
# Both pipes operators uses whats on the left side and uses it as the first argument in the next function (dependencies)


exscidata::dxadata |>
  lm(weight ~ height, data = _) # Low line means that data is fetched from what is left from the operator

exscidata::dxadata %>% 
  lm(weight ~ height, data = .) # The lm function cannot be piped further

# "Pipable" functions have "data" as the first argument

  
dplyr::select() # The select-function is in several packages, so smart to define that I want to use the one in the dplyr package


```

### Reading data into R

readr provides functions for reading and writing delimiter separated files, such as .csv

readxl provides functions for reading excel-files

### The verbs of data wrangling

Verb = take data and make different observations (in dplyr in the tidyverse package)

Useful functions select, rename, relocate, mutate, filter, arrange, summarize

chr = text data dbl = numeric data fctr =

#### The select function

```{r}

library(exscidata)
glimpse(exscidata::dxadata) # Function that shows us a glimpse of the data

exscidata::dxadata |>
  select(participant:weight, starts_with("lean.") & contains("_leg")) 
# : to choose every variable from x to y
# use start_with and/or contains to select variables


exscidata::dxadata |>
  select(weight, fp = participant) # Selecting weight, and selecting participant and change the name from participant to fp

exscidata::dxadata |>
  select(participant:weight) |> # Choose variables from participant to weight
  rename (kg = weight) |> # Rename the variable
  relocate(kg, .before = participant) # Relocate the order of variables (can use .after and . before)

# Type ?select in the console to get information about the function

exscidata::dxadata |>
  select(where(is.numeric)) # To check if variables are numeric


```

#### The mutate fuction

```{r}
library(exscidata)

exscidata::dxadata %>% 
  select(participant:weight, lean.whole) %>%   
  mutate(rel_lean = 100 * (lean.whole/1000/weight),# % lean mass of total body mass
         anewvariable = rnorm(80, 0, 1)) %>% 
  relocate(c(rel_lean, anewvariable), .before = participant) %>% # Combines variables by using c() and then relocate
  print()


```

### Tidy data

One variable is one coloumn, one observation is one row

#### Pivot data

```{r}

library(exscidata)

## Create a small dataset containing volume/leg information
leg_volume <- exscidata::dxadata %>% 
  select(participant, time, multiple, single) %>% 
  pivot_longer(names_to = "volume",  # Making a new column named "volume"
               values_to = "leg",    # Making a new column named "leg"
               cols = single:multiple) %>% 
  distinct(participant, volume, leg) %>% # Only keeping the unique combinations of variables, ie.the variables only appears once (not pre single + multiple and post single + multiple, but just single + multiple for each participant. Ie, each participant appears two times, not four)
  print()


## Create another small dataset containining lean mass/leg information
lean_mass <- exscidata::dxadata %>% 
  select(participant, time, starts_with("lean.") & contains("leg")) %>%
  select(-lean.legs) %>% # Write "contains ("_legs") instead to make code shorter
  pivot_longer(names_to = "leg",
               values_to = "leanmass",
               cols = starts_with("lean")) %>% 
  mutate(leg = if_else(leg == "lean.left_leg", "L", "R")) %>% # If leg is "lean.left_leg", change name to "L", if not change name to "R"
  print()


## Joining datasets

# Can use left_join, right_join, inner_join, full_join, anti_join
# Inner_join only uses variables that exist in both datasets (removes variables) so that we dont get missing values
# If your dataset is complete, it is not 

full_join(leg_volume, lean_mass)




```
